Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-03-06 mop/draft.tex
Updates from Overleaf

\documentclass{article}
\usepackage[margin=2cm]{geometry}
\usepackage{natbib}

\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}

@@

Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{doucet2011sf} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates. Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

@@

This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{doucet2011sf}) when $\alpha=1$. When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.

@@

We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance. This leads us to our first assumption.

\begin{aspt}[Smooth Neighborhood]
\label{assump:smooth-nbhd}
There exists a neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, is twice differentiable in $\theta$. 
\end{aspt}

\paragraph{Intuition:} We can justify this assumption as follows. For suitably nearby $\theta$ and $\phi$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same, eliminating any discontinuities from resampling, and that for small enough $\mathcal{N}(\phi)$ the likelihood ratios are bounded. The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are.

@@

\begin{aspt}[Bounded Measurement Model] There is an $\epsilon>0$ with $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n{ }^* \mid x_n; \theta\right)>\epsilon$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
\end{aspt}

\begin{aspt}[Locally Bounded Derivative]
\label{assump:local-bounded-derivative}
Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
    $$
    \|\nabla \ell(\theta, \phi, \omega, J)\|_2<G(\theta) \leq G<\infty
    $$
    for every $\phi \in \mathcal{M}, \theta$ in smooth neighborhood $\mathcal{N}(\phi)$, and almost every $\omega \in \Omega$.
\end{aspt}

@@

We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi$. The proof can be found in Appendix \ref{app:dop}.

@@

    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
    \begin{equation}
        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
    \end{equation}
\end{prop}

\kevin{Cut down on DOP discussion in actual manuscript, remove any references to this as an algorithm, just talk about the idea of an off-policy system/particle filter. Bottom paragraph is the punchline.}

@@

It is only ever practical to evaluate $\theta$ at $\phi$. This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate. Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}

@@

@@

The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney2016subsampled}. We defer the proof to Appendix \ref{app:ifad}

@@

Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
To see  (\ref{L2}),
we write the numerator of (\ref{lem:change-measure-proper-weights}) as
\[
L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
\]
Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
\[
L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
\]
from which we obtain (\ref{L2}).

@@

\begin{lem}
    For $\theta$, $\phi$ that satisfy Assumption \ref{assump:smooth-nbhd}, the following correction in the particle filter weights,
    \begin{equation}
        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
    \end{equation}
    successfully recovers $\lik(\theta)$.
\end{lem}

@@

\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$. Also let $\theta \in \gN(\phi)$. When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
    \begin{equation}
        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
    \end{equation}
\end{lem}

@@

\begin{lem}
    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then

@@

$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}.$$

Using the log-derivative identity again,
$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,k_j}^\theta - \log g_{i,k_j}^\phi\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,k_j}^\theta.$$

So 
$$  \nabla_\theta \sum_{n=1}^N \log g_{n,k_j}^\theta = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,k_j}^\theta\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,k_j}^{P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$

and 
$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{i=1}^N \nabla_\theta \log g_{n,k_j}^\theta = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$

@@

@@

@@

>>>>

\documentclass{article}
\usepackage[margin=2cm]{geometry}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}

@@

Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{poyiadjis11} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates. Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

@@

This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$. When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.

@@

@@

\begin{aspt}[Bounded Measurement Model]
\label{assump:bounded-measurement}
There is an $\epsilon>0$ such that $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)>\epsilon$ and a $G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty < G'(\theta)$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
\end{aspt}

\begin{aspt}[Locally Bounded Derivatives]
\label{assump:local-bounded-derivative}
Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
    $$
    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty <G(\theta) \leq G<\infty, \;\;
    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty <G(\theta) \leq G<\infty
    $$
    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
\end{aspt}

We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
    \label{assump:diff-meas}
    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$. 
\end{aspt}

In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
\begin{aspt}[Bounded And Differentiable Process Model]
    \label{assump:diff-process}
    Either one of the following two conditions hold:
    \begin{enumerate}
        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
    \end{enumerate}
\end{aspt}

We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.

We require many of these assumptions to establish the following proposition, the proof of which is a straightforward consequence of what we assumed.

\begin{prop}[Smooth Neighborhood]
\label{prop:smooth-nbhd}
There exists an open neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, has two continuous derivatives in $\theta$.
\end{prop}

\begin{proof}
    For a fixed $\omega$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same as the resampling indices in both cases correspond to that yielded by $f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}(\omega);\phi)$, eliminating any discontinuities from resampling.

    By Assumptions \ref{assump:diff-meas} and \ref{assump:diff-process}, for small enough $\mathcal{N}(\phi)$ the density ratios are bounded. The likelihood only changes by a factor of the density ratios, which are bounded and possess two continuous derivatives in $\theta$ if either the process and measurement densities used in the calculation do, or if the measurement densities do and one has access to a differentiable simulator with two continuous derivatives in $\theta$. This gives us the desired result.

    As a remark, we note that the neighborhoods the sequence is differentiable in are independent of $J$, as the resampling does not change with $\theta$ -- it is dependent on $\phi$. 
\end{proof}

@@

We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{poyiadjis11, scibior2021dpf} when $\theta$ is evaluated at $\phi$. The proof can be found in Appendix \ref{app:dop}.

@@

    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
    \begin{equation}
        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
    \end{equation}
\end{prop}

@@

It is only ever practical to evaluate $\theta$ at $\phi$. This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate. Its gradient in this case is either the estimate of \citet{poyiadjis11} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}

@@

\kevin{I think we can establish the linear convergence for the entire procedure in the same sense as \href{this}{https://arxiv.org/pdf/2311.08442.pdf}, in the sense that if IF2 is run for a sufficiently large but fixed number of iterations, once we arrive in a local strongly convex region, we get linear convergence when we swap to MOP. Show by induction that we remain in the strongly convex region (in terms of parameter space and likelihood), eqn 218. }

@@

The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney16}. We defer the proof to Appendix \ref{app:ifad}

@@

Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
To see  (\ref{L2}),
we write the numerator of (\ref{lem:change-measure-proper-weights}) as
\[
L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
\]
Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
\[
L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
\]
from which we obtain (\ref{L2}).

@@

\begin{lem}
    For $\theta$, $\phi$ that satisfy Proposition \ref{prop:smooth-nbhd}, the following correction in the particle filter weights,
    \begin{equation}
        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
    \end{equation}
    successfully recovers $\lik(\theta)$.
\end{lem}

@@

\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$. Also let $\theta \in \gN(\phi)$. When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
    \begin{equation}
        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
    \end{equation}
\end{lem}

@@

\begin{lem}
    \label{lem:mop-1-formula}
    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then

@@

$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}},$$
where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th particle at timestep $n$.

Using the log-derivative identity again,
$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,\theta} - \log g_{i,j}^{A,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,\theta}.$$

So 
$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$

and 
$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$

@@

\begin{lem}
\label{lem:mop-0-formula}

@@

\kevin{I have no clue how to prove this. Help!}
\begin{conj}[MOP-1 Estimate Is Unbiased for the Likelihood and the Gradient of the Likelihood]
    Consider the case of MOP-$\alpha$ when $\alpha=1$, and resampling is carried out with either multinomial, residual, or systematic resampling at every timestep. Assume the gradient estimate obtained by the gradient of the likelihood estimate with J particles, $\nabla_\theta \hat\lik_J(\theta)$ is uniformly bounded for all $J$, and that the likelihood has two continuous derivatives. Then the likelihood estimate is unbiased for the likelihood and the gradient of the likelihood estimate is unbiased for the gradient of the likelihood. That is, it holds that
    $$\E[\hat\lik_J(\theta)] = \lik(\theta),\;\; \E[\nabla_\theta \hat\lik_J(\theta)] = \nabla_\theta \lik(\theta).$$
\end{conj}

\begin{proof}
    First, we need to establish the unbiasedness of the likelihood estimate. This is 
    $$\hat\lik_J(\theta) := \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right)\prod_{n=1}^N L_n^\phi = \left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\prod_{n=1}^N L_n^\phi.$$

    We know from \cite{chopin20} that $\E[L_n^{\phi}] = \lik(\phi).$ It then suffices to show that
    $$\E\left[\left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\right] = \frac{\lik(\theta)}{\lik(\phi)}.$$

    To obtain this, we need to see that the resampling under $\phi$, with corrected weights, is unbiased. That is, we need to show that for any bounded measurable functional $\varphi : \gX \to \R$,
    $$\E\left[\frac{\sum_{j=1}^J w_{n,j}^{F, \theta} \varphi(x_{n,j}^{\theta, F})}{\sum_{j=1}^J w_{n,j}^{F, \theta}} \bigg| \gF_n \right] = \frac{\sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta})}{\sum_{j=1}^J w_{n,j}^{P, \theta} }. $$

    To see this, 
    \begin{align*}
        \sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta}) 
        &= \sum_{j=1}^J w_{n-1,j}^{F, \theta}\frac{g_{n,j}^{\theta}}{g_{n,j}^{\phi}}  \varphi(x_{n,j}^{P, \theta})
    \end{align*}

%     Observe that our differentiable simulator $\process(\cdot, \theta)$ and choice of weights now satisfy:

%     - A Feynman-Kac model $\left\{M_t, G_t\right\}$ such that:

% - the weight function $G_t$ may be evaluated pointwise (for all $t$ );

% - it is possible to simulate from $\mathbb{M}_0\left(\mathrm{~d} x_0\right)$ and from $M_t\left(x_{t-1}, \mathrm{~d} x_t\right)$ (for any $x_{t-1}$ and $t$ ).

% - The number of particles $N$.

% - The choice of an unbiased resampling scheme (e.g. multinomial, residual, stratified, systematic), that is, an algorithm to generate variables $A_t^{1: N}$ in $1: N$ such that $\mathbb{E}\left[\sum_{m=1}^N \mathbbm{1}\left\{A_t^m=n\right\}\right]=N W_{t-1}^n$; see previous chapter for more details.
\end{proof}

\begin{lem}[Consistency of MOP-1 Gradient Estimate]
    Consider the case of MOP-$\alpha$ when $\alpha=1$. Assume the gradient estimate obtained by the gradient of the log-likelihood estimate with J particles, $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ is uniformly bounded by $G(\theta)$ for all $J$, and that the likelihood has two continuous derivatives. Then the gradient estimate, $\nabla_\theta \hat\ell_J(\theta)$, is consistent for the score.
    \label{lem:mop-grad-consistency}
\end{lem}
\begin{proof}

    % We know that the variance of the MOP-1 gradient estimate $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ goes to zero coordinate-wise as $J\to\infty$ by the particle filter law of large numbers for bounded functionals of particles (\kevin{Prove general LLN for functionals of MOP under boundedness assumptions in L2, but maybe Hoeffding may be enough to show this, because we only want convergence in probability, we just need to show the expectation converges}), so it suffices to show that it is asymptotically unbiased.

    % Write $\hat\pi_{n,J}$ for the filtering distribution at timestep $n$ and with $J$ particles, $\hat\pi_{1:N,J}$ for the sequence of filtering distributions from timesteps $1$ to $N$, and $\pi_n, \pi_{1:N}$ for the respective posteriors.

    % From Lemma \ref{lem:lik-proper-weight}, we know that $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1:n-1}^*, \theta)$ is consistently estimated by the conditional likelihood estimate obtained by MOP-1, for all $\theta$ in an open neighborhood of $\phi$. The likelihood is simply the product of the conditional likelihoods, and similarly for the estimates, so the likelihood $\lik(\theta)$ is also consistently estimated by the likelihood estimate obtained by MOP-1, $\hat\lik_J(\theta)$, for all $\theta$ in an open neighborhood of $\phi$, $\gN(\phi)$.

    % We therefore have that 
    % $$\lim_{J\to\infty} \E[\hat\lik_J(\theta)] = \lik(\theta)$$
    % pointwise for all $\theta \in \gN(\phi)$. Given that the gradient is bounded uniformly for all $J$ and consistency for the likelihood estimate holds for all $\theta$ in an open neighborhood of $\phi$ (and therefore all $\theta'$ in a sufficiently small neighborhood of $\theta$), the bounded convergence theorem allows us to switch the expectation and derivative:
    % $$\nabla_\theta \E[\hat\lik_J(\theta)] 
    % = \E[\nabla_\theta \hat\lik_J(\theta)].$$

    % Take limits on both sides to find that 
    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
    % = \lim_{J \to \infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$

    % We then need to swap the limit and derivative on the left-hand side. This requires uniform convergence of $\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]$, i.e. that $$\sup_{\theta \in \gN(\phi)} ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} \to 0 \text{ as } J \to \infty,$$
    % or that $\forall \epsilon>0$ $\exists N\in \mathbb{N}$ such that $n>N => ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} < \epsilon$ $\forall \theta \in \gN(\phi).$ However, it is easier to instead use an indirect argument.

    % We then need to swap the limit and derivative on the left-hand side. This requires uniform convergence.

    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at. The sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}. The sequence is also uniformly equicontinuous. To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J(\theta, \omega)|_{\theta=\theta'}$ is also bounded by $H$ for almost every $\omega\in \Omega$ and every $\theta' \in \gN(\phi)$, including $\theta=\phi$. A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous. So the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$. 
    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
    $$||\nabla_\theta \hat\lik_J(\theta, \omega)-\nabla_\theta \hat\lik_J(\theta', \omega)||_\infty < \epsilon.$$

    \kevin{Our previous understanding was that the target is differentiable, and the sequence is differentiable in increasingly small open neighborhoods. This may not be true. The neighborhoods the sequence is differentiable in should be independent of $J$, as the resampling does not change with $\theta$ as it is dependent on $\phi$. Either way, the target is in each of the neighborhoods, and this may be sufficient in itself.}

    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence. We claim that there is only one subsequential limit. We know that when as the gradient is bounded by Assumption \ref{assump:local-bounded-derivative} we can treat the gradient as a bounded functional. So by Lemma \ref{lem:posterior-proper-weights} the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit. The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$, 
    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega).$$

    From Lemma \ref{lem:lik-proper-weight}, we know that
    $$\hat{\lik}_J(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta).$$

    We then have that for almost every $\omega \in \Omega$, 
    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega) = \nabla_\theta \lik(\theta),$$
    as we wanted. The result then follows by the continuous mapping theorem.

    % Taking expectations on both sides and using the bounded convergence theorem to interchange limits and expectations, we see that
    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] = \E[\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta)] = \E[\nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta)] = \nabla_\theta \lim_{J \to \infty} \E[\hat\lik_J(\theta)].$$

    % It then follows that
    % $$\nabla_\theta\lik(\theta) 
    % = \nabla_\theta \lim_{J\to\infty} \E[\hat\lik_J(\theta)] 
    % = \lim_{J\to\infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
    % = \lim_{J\to\infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$

    % We then have mean-square consistency for the gradient of the likelihood estimate. The result then follows by the continuous mapping theorem.

    \kevin{Explain that the difference in formula between this and Poyiadjis is due to us taking the derivative w.r.t. the particles in the chain rule. }

    \kevin{Alternative argument: Show gradient is unbiased, after showing the likelihood is unbiased in an open neighborhood of phi. That should hold by Chopin's argument for unbiased resampling to nonconstant weights. Gradient unbiased if can swap expectation and derivative, I think with two continuous derivatives and boundedness, might need BCT/DCT. No need to swap limit and derivative in this case, or worrying about uniform limits.}

@@

\section{Analysis of MOP-$\alpha$ Variance}

We first begin with a lemma based on the work of \cite{yu94}. \cite{kausik23} simplify this to the special case of Markov chains. We further adapt this to our setting below.
\begin{lem}
    Let $h: \gX^N \to \R^n$ be a bounded measurable function with bound $C$ on trajectories of points in the state space $\gX$ (by considering the domain $\gX^N$, this also encompasses functions of segments of trajectories). Further assume that the underlying Markov chain $(X_n)_{n=1}^N$ has mixing time $t_{\text {mix}}=t_{\text{mix}}\left(\frac{1}{4}\right)$. Consider the joint distribution $\chi$ over the product of the $\sigma$-algebras of $S$ such segments separated by $s$ timesteps, with marginals $\chi_i$. Let the product distribution of the marginals $\chi_i$ be called $\Xi$. Then, we have
$$
\left|\mathbb{E}_\chi h-\mathbb{E}_{\Xi} h\right| \leq \frac{4 C(S-1)}{4^{s\cdot t_{\text{mix}}}}.
$$

\end{lem}
\begin{proof}
    Remember that each of our Markov processes is mixing, so there exists $t_{m i x, j}=t_{m i x, j}\left(\frac{1}{4}\right)$ and a stationary distribution $d_j$ so that $T V\left(P_j^n(x, \cdot), d_j\right)<\frac{1}{4}$ for $n \geq t_{m i x, j}$. Let $t_{m i x}=\max _j t_{m i x, j}$. Since the decay in total variation distance is multiplicative, $T V\left(P_j^n(x, \cdot), d_j\right)<\left(\frac{1}{4}\right)^c$ for all $j$ and $n \geq c t_{m i x}$. This implies that
    $$
    \max _j T V\left(P_j^n(x, \cdot), d_j\right)<\left(\frac{1}{4}\right)^{\frac{T_n}{4 t_{m i x}}-1}=4 \lambda^n
    $$
    where $\lambda=\left(\frac{1}{4}\right)^{\frac{1}{t_{m i x}}}$.
\end{proof}

It is known from \cite{poyiadjis11} that the variance of MOP-1 is no smaller than $O(N^2)$, with an analogous upper bound on the MSE. We now address the case of $\alpha<1$.

We can decompose the MOP-$\alpha$ estimator as follows. When $\theta=\phi$,
$$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \phi}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \phi})^\alpha}.$$
